---
title: "03-compare-patch-protocols"
author: "Ariel Marcy"
date: "10/15/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../aus-rodent-skulls')
```

# Compare big and small patch protocols
Our study asked a methodological question: does it make a difference if we use big areas of patch semi-landmarks versus breaking it up with curve semilandmarks? Or in other words, will a big patch landmarks pick up different biological patterns than an area broken up by curve landmarks defined along sutures?

We hypothesized that the big patch protocol would pick up convergent evolution while the small patch + suture curve protocol would better detect phylogentically-conserved shapes. 

Here we test if the shape patterns produced by the two protocols are different in any way.

### Load packages, functions, and data from previous steps
```{r message = FALSE}
library(data.table)
library(geomorph)
source("../Functions/utilities.R")  # custom functions
load(file = "../Data/Processed/02-data-for-analyses.rda")

# Load libraries (and install if necessary) for landvR package
if(!require(devtools)) install.packages("devtools")
if(!require(dispRity)) install.packages("dispRity")
if(!require(landvR)) install_github("TGuillerme/landvR")
```

# Visualizing landmark variation
Visualizing landmarks is important both to detect landmarking errors as well as to better understand which parts of the specimen vary the most. 

We investigate both in this script. In fact, the new landvR package by Dr Thomas Guillerme used _geomorph_'s `plotRefToTarget()` function as its base to visualize landmark variation with a heatmap.

## Landmark variation heatmaps
Assuming that all landmarks are correctly placed now, we can use a variation of the `procrustes.var.plot()` function by Dr Thomas Guillerme to see how landmarks vary within and between different taxa. 

Much of the below is adapted from the [vignette written by Dr Guillerme here](https://cdn.rawgit.com/TGuillerme/landvR/8a6a6bd5/inst/vignettes/Landmark_partition_test.html).
```{r}
# Find mean configuration
sm.consensus <- (select.procrustes(sm.shape, selector = mean))[[1]]

# Measure spherical coordinates differences from the mean
diff.from.mean <- coordinates.difference(coordinates = sm.shape, reference = sm.consensus, type = "spherical")
```

To test the hypothesis that each landmark's variation from the mean configuration is above expected, we can use the PC axes from a PCA.
```{r}
# Ordinate the data
twoD.shape <- two.d.array(sm.shape)
ordination <- stats::prcomp(twoD.shape)

# Force the symmetric component of shape into class "gpagen" (required for variation.range)
sm.gpagen.shape <- list()
sm.gpagen.shape$coords <- sm.shape
sm.gpagen.shape$consensus <- sm.consensus
class(sm.gpagen.shape) <- "gpagen"

# Measure extremes of variation from mean on PC1
PC1.var <- variation.range(sm.gpagen.shape, return.ID = FALSE, axis = 1, ordination = ordination, type = "spherical")
```

The next step is to make a heatmap showing the size of variation and direction of movement for each landmark in the dataset, comparing between the min and max specimens along PC1.
```{r}
# Wrap specimens on the tangent space
wrap.PCA <- plotTangentSpace(sm.shape, verbose = FALSE)

# Select extreme specimens (the ones that make the warp-meshes)
hypothetical.1 <- wrap.PCA$pc.shapes[[1]]
hypothetical.2 <- wrap.PCA$pc.shapes[[2]]

# Plot the range of variation along PC1 using a heat color scheme
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = c("yellow", "orange", "red"), col.val = PC1.var[, "radius"], labels = F)

# Plot the histogram of variation in landmarks for this protocol
var.range <- variation.range(sm.gpagen.shape)
hist(var.range[, 1])
```

## REPEAT FOR BIG PATCH PROTOCOL

## Plot PCAs for each protocol
PCAs allow visual comparison of how differently the patch protocols sort the shape data. To assist, we wrote a function, `PlotByGroup()`, to match colors or other plotting attributes to groups within the dataset.

Here, we'll use this function to look at the morphospace colored by genus.
```{r}
# Colors by genus: Hyd Leg Lep Mas Mel Mes Mus Not Pog Pse Rat Uro Xer Zyz
col.gen <- c("red", "cornflowerblue", "light green", "dark blue", "yellow", "green", "grey", "blue", "magenta", "dark blue", "black", "darkgoldenrod4", "orange", "light blue")

sp.col.gen <- PlotByGroup(big.info, "Genus", col.gen)  # is the same for big and small protocols, tested with all.equal()
```

### PCA plot of big patch protocol
PCAs are very human-friendly way of visualizing shape variation in 2 dimensions. Usually, the proportion of variance for each PC axis is reported in percent. 
```{r}
big.pca <- plotTangentSpace(big.shape, groups = sp.col.gen, axis1 = 1, axis2 = 2, verbose = T)

# Get proportion of variance for PC1 and PC2 in format for plots
big.PCs <- big.pca$pc.summary$importance
big.PC1 <- round(big.PCs[2, 1] * 100, digits = 1)  # % with 1 decimal
big.PC2 <- round(big.PCs[2, 2] * 100, digits = 1)

# Pretty plot
plot(x = big.pca$pc.scores[, 1], 
     y = big.pca$pc.scores[, 2], 
     xlab = paste("PC1 (", big.PC1, "%)", sep = ""), 
     ylab = paste("PC2 (", big.PC2, "%)", sep = ""), 
     asp = TRUE, 
     col = sp.col.gen, 
     bg = sp.col.gen,
     pch = 16, 
     cex = 1.5, 
     cex.axis = 1.3, 
     cex.lab = 1.3)
```

### PCA plot of small patch protocol
```{r}
# Plot PCAs
sm.pca <- plotTangentSpace(sm.shape, groups = sp.col.gen, axis1 = 1, axis2 = 2, verbose = T)

# Get proportion of variance for PC1 and PC2 in format for plots
sm.PCs <- sm.pca$pc.summary$importance
sm.PC1 <- round(sm.PCs[2, 1] * 100, digits = 1)
sm.PC2 <- round(sm.PCs[2, 2] * 100, digits = 1)

# Pretty plot
plot(x = sm.pca$pc.scores[, 1], 
     y = sm.pca$pc.scores[, 2], 
     xlab = paste("PC1 (", sm.PC1, "%)", sep = ""), 
     ylab = paste("PC2 (", sm.PC2, "%)", sep = ""), 
     asp = T,
     col = sp.col.gen, 
     bg = sp.col.gen,
     pch = 16, 
     cex = 1.5, 
     cex.axis = 1.3, 
     cex.lab = 1.3)
```

## Compare performance of patch protocols in ANOVA by genus and centroid size
The patch protocols look very similar on the PCAs, which suggests that they do not capture shape variation differently from one another. To test this, we performed two different Procrustes ANOVAs on each protocol's dataset:
1) shape ~ genus 
2) shape ~ centroid size (a proxy for body size), which is a test for allometry

### Run shape ~ genus * Csize ANOVAs on each protocol:  
Big patch protocol:
```{r}
gdf <- geomorph.data.frame(genus = big.info$Genus, size = big.info$Csize)
big.procD <- procD.lm(big.shape ~ genus * size, data = gdf)
print(big.procD)
```

Small patch protocol:
```{r}
gdf <- geomorph.data.frame(genus = sm.info$Genus, size = sm.info$Csize)
sm.procD <- procD.lm(sm.shape ~ genus * size, data = gdf)
print(sm.procD)
```

These ANOVAs are also very similar. Therefore, we decided to use only the small patch protocol as this protocol captured more information (sutures versus patch points) and had less asymmetry than the big patch protocol. It also resembles the more common method of landmarking.

### Save intermediate data
```{r}
# Remove "sm." from shape and metadata matrix names for increased readability
shape <- sm.shape
info <- sm.info
PCs <- sm.PCs
col.gen <- sp.col.gen

save(shape, info, PCs, col.gen, file = "../Data/Results/05-sm-data-for-analyses.rda")
```
