---
title: "05-compare-patch-protocols"
author: "Ariel Marcy"
date: "10/15/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../aus-rodent-skulls')
```

# Compare big and small patch protocols
Our study asked a methodological question: does it make a difference if we use big areas of patch semi-landmarks versus breaking it up with curve semilandmarks? Or in other words, will a big patch landmarks pick up different biological patterns than an area broken up by curve landmarks defined along sutures?

We hypothesized that the big patch protocol would pick up convergent evolution while the small patch + suture curve protocol would better detect phylogentically-conserved shapes. 

Here we test if the shape patterns produced by the two protocols are different in any way.

### Load packages and functions
```{r message = FALSE}
library(data.table)
library(geomorph)
source("../Functions/utilities.R")  # loads custom functions called by these scripts but defined in the Functions/utilities.R file for conciseness.
```

### Load data from previous steps
```{r}
load(file = "../Data/Results/01-big-patch-data.rda")
load(file = "../Data/Results/01-sm-patch-data.rda")
load(file = "../Data/Results/02-GPA-data.rda")
load(file = "../Data/Results/04-data-for-analyses.rda")

```

## Plot PCAs for each protocol
PCAs allow visual comparison of how differently the patch protocols sort the shape data. To assist, we wrote a function, `PlotByGroup()`, to match colors or other plotting attributes to groups within the dataset.

Here, we'll use this function to look at the morphospace colored by genus.
```{r}
# Colors by genus: Hyd Leg Lep Mas Mel Mes Mus Not Pog Pse Rat Uro Xer Zyz
col.gen <- c("red", "cornflowerblue", "light green", "dark blue", "yellow", "green", "grey", "blue", "magenta", "dark blue", "black", "darkgoldenrod4", "orange", "light blue")

sp.col.gen <- PlotByGroup(big.info, "Genus", col.gen)  # is the same for big and small protocols, tested with all.equal()
```

### PCA plot of big patch protocol
PCAs are very human-friendly way of visualizing shape variation in 2 dimensions. Usually, the proportion of variance for each PC axis is reported in percent. 
```{r}
big.pca <- plotTangentSpace(big.shape, groups = sp.col.gen, axis1 = 1, axis2 = 2, verbose = T)

# Get proportion of variance for PC1 and PC2 in format for plots
big.PCs <- big.pca$pc.summary$importance
big.PC1 <- round(big.PCs[2, 1] * 100, digits = 1)  # % with 1 decimal
big.PC2 <- round(big.PCs[2, 2] * 100, digits = 1)

# Pretty plot
plot(x = big.pca$pc.scores[, 1], 
     y = big.pca$pc.scores[, 2], 
     xlab = paste("PC1 (", big.PC1, "%)", sep = ""), 
     ylab = paste("PC2 (", big.PC2, "%)", sep = ""), 
     asp = TRUE, 
     col = sp.col.gen, 
     bg = sp.col.gen,
     pch = 16, 
     cex = 1.5, 
     cex.axis = 1.3, 
     cex.lab = 1.3)
```

### PCA plot of small patch protocol
```{r}
# Plot PCAs
sm.pca <- plotTangentSpace(sm.shape, groups = sp.col.gen, axis1 = 1, axis2 = 2, verbose = T)

# Get proportion of variance for PC1 and PC2 in format for plots
sm.PCs <- sm.pca$pc.summary$importance
sm.PC1 <- round(sm.PCs[2, 1] * 100, digits = 1)
sm.PC2 <- round(sm.PCs[2, 2] * 100, digits = 1)

# Pretty plot
plot(x = sm.pca$pc.scores[, 1], 
     y = sm.pca$pc.scores[, 2], 
     xlab = paste("PC1 (", sm.PC1, "%)", sep = ""), 
     ylab = paste("PC2 (", sm.PC2, "%)", sep = ""), 
     asp = T,
     col = sp.col.gen, 
     bg = sp.col.gen,
     pch = 16, 
     cex = 1.5, 
     cex.axis = 1.3, 
     cex.lab = 1.3)
```

## Compare performance of patch protocols in ANOVA by genus and centroid size
The patch protocols look very similar on the PCAs, which suggests that they do not capture shape variation differently from one another. To test this, we performed two different Procrustes ANOVAs on each protocol's dataset:
1) shape ~ genus 
2) shape ~ centroid size (a proxy for body size), which is a test for allometry

### Add centroid size from gpagen() to symmetric shape metadata
Since the bilat.symmetry() function changed the order of specimens, we need to add the centroid size information from gpagen() to the new metadata table we just made for the symmetric component of shape.
```{r}
# Remove specimen filenames from dimnames of centroid size vectors and combine CatNum with centroid data
big.CatSize <- as.data.table(cbind(as.vector(big.sp.info$CatNum), unname(big.Y$Csize)))  # data from GPA in script 02
names(big.CatSize) <- c("CatNum", "Csize")
big.CatSize$Csize <- as.numeric(big.CatSize$Csize)  # coerce back to numeric

sm.CatSize <- as.data.table(cbind(as.vector(sm.sp.info$CatNum), unname(sm.Y$Csize)))
names(sm.CatSize) <- c("CatNum", "Csize")
sm.CatSize$Csize <- as.numeric(sm.CatSize$Csize)

# Remove error replicates from centroid size table
big.CatSize <- big.CatSize[which(big.sp.info$Rep == "non"), ]
sm.CatSize <- sm.CatSize[which(sm.sp.info$Rep == "non"), ]

# Merge Csize with bilateral metadata for each patch protocol
big.info <- merge(big.info, big.CatSize, by = "CatNum", sort = F)
sm.info <- merge(sm.info, sm.CatSize, by = "CatNum", sort = F)
```

### Run shape ~ genus * Csize ANOVAs on each protocol:  
Big patch protocol:
```{r}
gdf <- geomorph.data.frame(genus = big.info$Genus, size = big.info$Csize)
big.procD <- procD.lm(big.shape ~ genus * size, data = gdf)
print(big.procD)
```

Small patch protocol:
```{r}
gdf <- geomorph.data.frame(genus = sm.info$Genus, size = sm.info$Csize)
sm.procD <- procD.lm(sm.shape ~ genus * size, data = gdf)
print(sm.procD)
```

These ANOVAs are also very similar. Therefore, we decided to use only the small patch protocol as this protocol captured more information (sutures versus patch points) and had less asymmetry than the big patch protocol. It also resembles the more common method of landmarking.

### Save intermediate data
```{r}
# Remove "sm." from shape and metadata matrix names for increased readability
shape <- sm.shape
info <- sm.info
PCs <- sm.PCs
col.gen <- sp.col.gen

save(shape, info, PCs, col.gen, file = "../Data/Results/05-sm-data-for-analyses.rda")
```