---
title: "06-analyze-shape-with-phylogeny"
author: "Ariel Marcy"
date: "11/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../aus-rodent-skulls')
```

# Phylogenetic analyses
Evolutionary studies must take into account the phylogenetic tree because related species do not behave the same as independent samples. 

### Load packages, functions, and data
```{r message = FALSE}
library(ape)  # needed to load in .tre phylogeny files
library(stringr)
library(data.table)
library(geomorph)
source("../Functions/utilities.R")  # loads custom functions
load(file = "../Data/Processed/03-sm-data-for-analyses.rda")
load(file = "../Data/Processed/04-sm-data-w-traits.rda")
```

### Load in phylogenetic tree
The most recent tree for Australian rodents is from Smissen & Rowe 2018, Figure 2. The tree is a fossil calibrated ultrametric tree from BEAST2 for Hydromini, which includes all the Australo-Papuan old endemics in Murinae.
```{r}
aus.rodents <- "../Data/Processed/Smissen-Rowe-2018-concat.tre"  # 
raw.tree <- read.tree(aus.rodents)
```

## 

