---
title: "07-plot-exploratory-PCAs"
author: "Ariel Marcy"
date: "10/17/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../aus-rodent-skulls')
```

# Plotting exploratory PCAs
PCAs provide a human-friendly way to visualize shape data. They are a good way to understand what's going on in the data quickly. However, because they artificially collapse the shape variation into orthogonal PC axes, they must be considered exploratory analyses only.

### Load packages
```{r message = FALSE}
library(stringr)
library(data.table)
library(geomorph)
```

### Load data from previous steps
```{r}
load(file = "../Data/Results/05-sm-data-for-analyses.rda")
```

## Exploratory analyses with PCAs
Here, we explore the variation in rodent skulls by coloring the PCA plots by genus, taxa, and by traits.

Since we are going to add several columns to the metadata table, we'll remove columns no longer needed past the bilateral symmetry and replicate handling stages. 
```{r}
# Prep metadata for column additions
info <- info[, c(-4,-6)]  # removes unnecessary "Rep" and "All" columns
info$Order <- c(1:dim(info)[1])  # preserves specimen order so metadata matches shape data
```

## Plot PCA by genus for small patch dataset with no landmarking errors
This plot will be similar to the small patch PCA plot from script 05 but using a dataset that's been checked for landmarking errors in script 06.

### Write function to make pretty PCA plots quickly
This function adapts code used in script 05.
```{r}
PlotPCA <- function(shape, PCx, PCy, col.grp, pch.grp = 16) {
        # Plots PCAs with specimens colored (and optionally given different points) according to groups, reports PC axis variation in %.
        # 
        # Args:
        #    shape: a 3D array of shape coordinates in (p x k x n) format
        #    PCx: the PC intended for the x-axis.
        #    PCy: the PC intended for the y-axis, usually x > y. 
        #    col.grp: a vector of colors ordered in the same way as specimens, usually made with PlotByGroup() function.
        #    pch.grp: an optional vector for point shapes, also usually made with PlotByGroup() function. Default is a filled circle.
        
        pca <- plotTangentSpace(shape, groups = col.grp, axis1 = PCx, axis2 = PCy, verbose = T)
        
        # Write x and y labels with proportion of variance for PCx and PCy
        PCs <- pca$pc.summary$importance
        PCx.per <- round(PCs[2, PCx] * 100, digits = 1)  # % with 1 decimal
        PCx.lab <- paste("PC", PCx, " (", PCx_per, "%)", sep = "")
        PCy.per <- round(PCs[2, PCy] * 100, digits = 1)
        PCy.lab <- paste("PC", PCy, " (", PCy_per, "%)", sep = "")
        
        PCA.plot <- plot(x = pca$pc.scores[, PCx],
                         y = pca$pc.scores[, PCy], 
                         xlab = PCx.lab, 
                         ylab = PCy.lab,
                         asp = TRUE,
                         col = col.grp, 
                         pch = pch.grp, 
                         bg = col.grp,
                         cex = 1.5,
                         cex.axis = 1.3, 
                         cex.lab = 1.3)
        }
```

The chunk below demonstrates how a plot can be exported as a .eps file. These file types can be opened in programs like Adobe Illustrator for minor aesthetic adjustments or be exported to a variety of other filetypes.
```{r}
# Plot PCA by genus and export as .eps into ../Data/Results folder
setEPS()
postscript("../Data/Results/PCA_genus.eps")
PlotPCA(shape, 1, 2, col.gen)
dev.off()
```

## PCA plot by diet and guild
This PCA plot colors specimen points by diet and gives them a point shape based on guild, i.e. where they spend most of their time foraging: aquatic, arboreal, or terrestrial.

Trait information was gathered from the book, Native Mice and Rats by Bill Breed and Fred Ford (2006). 
```{r}
# Add trait data to the metadata table, info and preserve specimen order
traits <- read.csv("../Data/Processed/in_ex_traits.csv", header = TRUE)
traits$Taxa <- paste(str_sub(traits$Genus, 1, 1), str_sub(traits$Species, 1, 3), sep = "_")  # make matching Taxa column
info.traits <- merge(info, traits[3:8], by = "Taxa", sort = F)  # adds only new columns from the trait data frame
info.traits <- info.traits[order(info.traits$Order),] 
```

Assign colors by diet and point shapes by guild for PCA.
```{r}
# Colors by Carnivorous, Frugivorous, Granivorous, Herbivorus, Omnivorous
col.feed <- c("red", "darkorange", "gold", "green", "burlywood4")
col.feed <- PlotByGroup(info.traits, "Feeding", col.feed)

# Point key: Aquatic (circle), Arboreal (triangle), Terrestrial (square)
pch.gld <- c(19, 24, 22)
pch.gld <- PlotByGroup(info.traits, "Guild", pch.gld)
```

Plot PC1 vs PC2 and add custom legends.
```{r}
PlotPCA(shape, 1, 2, col.feed, pch.gld)

# Legends
legend(0.071, -0.012, legend = c("Aquatic", "Arboreal", "Terrest."), col = "black", border = NULL, pch = c(21, 24, 22))
text(-0.0, -0.055, "Carnivore", col = "red")
text(0.083, -0.005, "Frugivore", col = "darkorange")
text(-0.015, 0.033, "Herbivore", col = "green")
text(-0.07, -0.045, "Omnivore", col = "burlywood4")
text(-0.085, 0.01, "Granivore", col = "gold2")
```

```{r}
save(info.traits, col.feed, pch.gld, file = "../Data/Results/07-sm-data-w-traits.rda")
```