---
title: "07-plot-exploratory-PCAs"
author: "Ariel Marcy"
date: "10/17/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../aus-rodent-skulls')
```

# Plotting exploratory PCAs
PCAs provide a human-friendly way to visualize shape data. They are a good way to understand what's going on in the data quickly. However, because they artificially collapse the shape variation into orthogonal PC axes, they must be considered exploratory analyses only.

### Load packages and functions
```{r message = FALSE}
library(stringr)
library(data.table)
library(geomorph)
source("../Functions/utilities.R")  # loads custom functions called by these scripts but defined in the Functions/utilities.R file for conciseness.
```

### Load data from previous steps
```{r}
load(file = "../Data/Results/05-sm-data-for-analyses.rda")
```

## Exploratory analyses with PCAs
Here, we explore the variation in rodent skulls by coloring the PCA plots by genus, taxa, and by traits.

Since we are going to add several columns to the metadata table, we'll remove columns no longer needed past the bilateral symmetry and replicate handling stages. 
```{r}
# Prep metadata for column additions
info <- info[, c(-4,-6)]  # removes unnecessary "Rep" and "All" columns
info$Order <- c(1:dim(info)[1])  # preserves specimen order so metadata matches shape data
```

## Plot PCA by genus for small patch dataset with no landmarking errors
This plot will be similar to the small patch PCA plot from script 05 but using a dataset that's been checked for landmarking errors in script 06.

### Write function to make pretty PCA plots quickly
The function `PlotPCA()` adapts code used in script 05 to quickly plot a PCA with specimens colored by groups and optionally given different point shapes for other attributes. The function also reports PC axis variation in %. 

The chunk below demonstrates how a plot can be exported as a .eps file. These file types can be opened in programs like Adobe Illustrator for minor aesthetic adjustments or be exported to a variety of other filetypes.
```{r}
# Plot PCA by genus and export as .eps into ../Data/Results folder
setEPS()
postscript("../Data/Results/PCA_genus.eps")
PlotPCA(shape, 1, 2, col.gen)
dev.off()
```

## PCA plot by diet and guild
This PCA plot colors specimen points by diet and gives them a point shape based on guild, i.e. where they spend most of their time foraging: aquatic, arboreal, or terrestrial.

Trait information was gathered from the book, Native Mice and Rats by Bill Breed and Fred Ford (2006). 
```{r}
# Add trait data to the metadata table, info and preserve specimen order
traits <- read.csv("../Data/Processed/in_ex_traits.csv", header = TRUE)
traits$Taxa <- paste(str_sub(traits$Genus, 1, 1), str_sub(traits$Species, 1, 3), sep = "_")  # make matching Taxa column
info.traits <- merge(info, traits[ , -c(1, 2)], by = "Taxa", sort = F)  # adds only new columns from the trait data frame
info.traits <- info.traits[order(info.traits$Order), ] 
```

Assign colors by diet and point shapes by guild for PCA.
```{r}
# Colors by Carnivorous, Frugivorous, Granivorous, Herbivorus, Omnivorous
col.feed <- c("red", "darkorange", "gold", "green", "burlywood4")
col.feed <- PlotByGroup(info.traits, "Feeding", col.feed)

# Point key: Aquatic (circle), Arboreal (triangle), Terrestrial (square)
pch.gld <- c(19, 24, 22)
pch.gld <- PlotByGroup(info.traits, "Guild", pch.gld)
```

Plot PC1 vs PC2 and add custom legends.
```{r}
PlotPCA(shape, 1, 2, col.feed, pch.gld)

# Legends
legend(0.071, -0.012, legend = c("Aquatic", "Arboreal", "Terrest."), col = "black", border = NULL, pch = c(21, 24, 22))
text(-0.0, -0.055, "Carnivore", col = "red")
text(0.083, -0.005, "Frugivore", col = "darkorange")
text(-0.015, 0.033, "Herbivore", col = "green")
text(-0.07, -0.045, "Omnivore", col = "burlywood4")
text(-0.085, 0.01, "Granivore", col = "gold2")
```

## Look for morphospace partitioning by habitat
So far, most of the variation in the morphospace could be explained by size and genus affiliation, with different genera of similar sizes overlapping quite a bit. This looks like there isn't much differentiation among Australian rodents, but perhaps the differentiation is hidden by geographical region or partitioning of the environment. 

### Write functions to subset and visualize species morphospace by region
First we will subset the PCA scores by species which inhabit the 7 different ecological regions of Australia. Regions and species distributions therein taken from Breed & Ford 2006. `FoundInRegion()` subsets existing PCA data by species present in given region(s).

Next, we wrote a function, `PointOutDiffSpp()` to give plotting points (pch values) by unique species within each genus.

Finally, we'll combine these functions with a loop to plot PCAs with color by taxa for each region, unique species within each genera having a unique point shape, and a legend for only the species present in each region. 
```{r}
# Get PCA data as used above
pca <- PlotPCA(shape, 1, 2, col.feed, pch.gld, return.PCA = TRUE)
pc.scores <- subset(pca$pc.scores, select = 1:3)  # only 1st 3 PCs

# Loop through all 7 regions to plot morphospace of species found in each
region.names <- c("Savannah", "Arid Zone", "NE Wet Forest", "NE Dry Forest", "Pilbara", "South West", "South East")

for (i in 1:7) {
        # Subset info.traits and PC scores by region
        in.region <- FoundInRegion(info.traits, regions = i)
        info.region <- info.traits[in.region, ]
        pc.region <- as.data.frame(pc.scores[in.region, ])
        
        # Colors by genus, points by unique species in a genus
        col.gen.region <- PlotByGroup(info.region, "Genus", col.gen)  # col.gen colors defined in script 05
        pch.tax.legend <- PointOutDiffSpp(info.region)
        pch.tax.region <- PlotByGroup(info.region, "Taxa", pch.tax.legend)
        
        # Plot with fancy (i.e. verbose) legend
        info.region$Legend <- paste(str_sub(info.region$Genus, 1, 3), str_sub(info.region$Species, 1, 3))  # Mus musculus -> "Mus mus"
        
        # Make color vector for legend points
        genera.in.region <- which(levels(info.traits$Genus) %in% sort(unique(info.region$Genus)))
        col.gen.reg <- col.gen[genera.in.region]  # subset col.gen by genera in region
        unique.taxa <- sort(unique(info.region$Legend))
        legend.genus <- substr(unique.taxa, 1, 3)
        names(col.gen.reg) <- unique(legend.genus)
        col.gen.legend <- col.gen.reg[match(legend.genus, names(col.gen.reg))]
        
        # Write x and y labels with proportion of variance for PCx and PCy
        PCs <- pca$pc.summary$importance
        PCx.per <- round(PCs[2, 1] * 100, digits = 1)  # % with 1 decimal
        PCx.lab <- paste("PC1 (", PCx.per, "%)", sep = "")
        PCy.per <- round(PCs[2, 2] * 100, digits = 1)
        PCy.lab <- paste("PC2 (", PCy.per, "%)", sep = "")
        
        PCA.plot <- plot(x = pc.region$PC1,
                         y = pc.region$PC2,
                         main = paste(region.names[i], "Region Morphospace" ),
                         xlab = PCx.lab,
                         ylab = PCy.lab,
                         xlim = c(-0.10, 0.095),
                         ylim = c(-0.04, 0.07),
                         col = col.gen.region, 
                         pch = pch.tax.region, 
                         bg = col.gen.region,
                         cex = 1.5,
                         cex.axis = 1.3, 
                         cex.lab = 1.3)
        legend(0.075, 0.075, legend = unique.taxa, col = col.gen.legend, border = NULL, pch = pch.tax.legend, pt.bg = col.gen.legend, cex = 0.64)
}
```

Plot PCAs (color by taxa) for forested regions by terrestrial vs arboreal
```{r}

```


### Save intermediate data
```{r}
save(info.traits, col.feed, pch.gld, file = "../Data/Results/07-sm-data-w-traits.rda")
```